Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-16 doc/thesis/conclusion.tex
Wrote discussion mostly

@@

Our work had three main aims. First, we developed a method to estimate contact
network model parameters from phylogenetic data. This method widens the field
of models which may be investigated in a phylodynamic context. Second, we
investigated the parameters of the \gls{BA} model. We determined through
simulation studies that the preferential attachment power $\alpha$ and number
of infected nodes $I$ had a substantial impact on transmission tree shape, and
could be estimated using our method.

@@

An alternative approach is the deterministic framework outlined by
\textcite{morris1993epidemiology}, who proposes to apply the standard
compartmental modelling framework to contact networks by assigning each
individual their own compartment. Thus, each individual is associated with a
single \gls{ODE}, with the entire \gls{ODE} system parameterized by the
adjacency matrix of the contact network. \citeauthor{morris1993epidemiology}
proposes to use log-linear models to parameterize the matrix. This framework is
highly expressive, and allows straightforward incorporation of time-dependent
dynamics. However, simulating a transmission tree would require the numerical
solution of a very large system of \glspl{ODE}. Given the large number of
simulations required for kernel-\gls{ABC}, it is not clear if this method would
be computationally feasible in this context.

% spatial networks, see section 5.4 in keeling2005networks

The two-step process of simulating a contact network and subsequently allowing
an epidemic to spread over that network carries with it the assumption that the
contact network is static over the duration of the epidemic. Clearly this
assumption is invalid, as people make and break partnerships on a regular
basis. Our work has not addressed this assumption, primarily due to our desire
to avoid the additional complexity required to address the dynamic nature of
networks. This simplifying assumption is made by most studies using contact
network models in an epidemiological context~\autocite{welch2011statistical,
bansal2007individual}. However, in principle, kernel-\gls{ABC} could be
adapted to dynamic contact networks by using a method such as that developed by
\textcite{robinson2012dynamics} to simulate a dynamic contact network, while
concurrently simulating the spread of an epidemic.

% Caveats

It is important to note that our kernel-ABC method takes a transmission tree as
input, rather than a viral phylogeny. Thus, we have left the estimation of a
transmission tree up to the user. There were two reasons for this choice. First
and foremost, we wished again to avoid extra complexity and keep the number of
estimated parameters small. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. Although this
may be an avenue for future extension, we felt that it would obscure the
primary purpose of this work, which is to study contact network parameters.
Second, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
jombart2011reconstructing, cottam2008integrating, poon2015phylodynamic}, some
of which incorporate geographic and/or epidemiological data not accommodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred tree building method.

>>>>

Due to the rapid advancement of nucleotide sequencing technology, viral
sequence data data have become increasingly feasible to collect on a population
level. Through phylodynamic methods, these data offer a window into
epidemiological processes which would otherwise be virtually impossible to
study on a realistic scale. Although phylodynamic methods have been developed
to fit a wide variety of epidemiological models to phylogenetic
data~\autocite{volz2012complex, rasmussen2014phylodynamic}, our method is the
first (to our knowledge) which can fit models that do not assume panmixia. We
believe this capability will be of broad interest to the molecular evolution
and epidemiology community, as it widens the field of epidemiological
parameters which may be investigated through viral sequence data. In addition,
the characterization of local contact networks could be valuable from a public
health perspective, such as for investigating optimal vaccination strategies.
This information could assist in curtailing current epimemics, as well as
preventing future epidemics of different diseases over the same contact
network.

@@

Our work had three main aims. First, we developed phylodynamic a method to
estimate contact network model parameters.

Our second aim was to investigate the parameters of the \gls{BA} model, in
terms of their impact on tree shape and our ability to reconstruct them with
kernel-\gls{ABC}. We determined through simulation studies that the
preferential attachment power $\alpha$ and number of infected nodes $I$ had a
substantial impact on transmission tree shape, and could be estimated using our
method.

The final aim of our work was to apply our method to real world
\gls{HIV} datasets.

@@

### 2016-04-16 doc/thesis/results.Rnw
Wrote discussion mostly

<<setup, include=FALSE>>=
    library(netabc)
@

\subsection{Kernel classifiers}

@@

@@

    alpha.sackin.cor <- alpha.stats[,cor.test(alpha, sackin, method="spearman")]
    stopifnot(alpha.sackin.cor$p.value < 1e-5)

    N.ratio.cor <- N.stats[,cor.test(int.tip.ratio, N)]
    I.ratio.cor <- I.stats[,cor.test(int.tip.ratio, I)]
    stopifnot(N.ratio.cor$p.value < 1e-5)
    stopifnot(I.ratio.cor$p.value < 1e-5)
@

Trees simulated under different values of \gls{alpha} are visibly quite
distinct (\cref{fig:alphatrees}). In particular, higher values of \gls{alpha}
produce networks with a small number of highly connected nodes which, once
infected, are likely to transmit to many other nodes. This results in a more
unbalanced, ladder-like structure in the phylogeny, compared to networks with
lower \gls{alpha} values. Sackin's index was significantly correlated with
\gls{alpha} 
    (Spearman's rho \Sexpr{round(alpha.sackin.cor$estimate, 2)},
     $p < 10^{-5}$).
None of the other three parameters produced trees which were as easily
distinguished from each other
(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}). However, the ratio of
internal to terminal branch lengths was positively correlated with $N$ and
negatively correlated with $I$ (Spearman's rho
    \Sexpr{round(N.ratio.cor$estimate, 2)} and
    \Sexpr{round(I.ratio.cor$estimate, 2)},
    both $p < 10^{-5}$).

@@

\subsection{Grid search}

The accuracy of grid search estimates largely paralleled that of the \gls{kSVM}
classifiers.

We used grid search to obtain marginal estimates for each network parameter
while holding all other parameters fixed. We observed that kernel scores were
highest at the values of \gls{alpha} on the grid closest to the true values, as
shown in Figure~\ref{fig:gridkernel}. However, there was a much stronger spike
in kernel scores near the true value for \gls{alpha} = 1.0 and 1.25. This is
recapitulated when we look at the accuracy of point estimates obtained by
taking the grid value with the highest median kernel score. As shown in
Figure~\ref{fig:gridest}, while the estimates are generally close to the true
value, they are much closer for \gls{alpha} = 1.25 than for the other values.

\begin{figure}[ht]
  \centering
  \includegraphics{gridsearch-alpha-kernel.pdf}
  \caption[Grid search kernel scores]{
    Grid search kernel scores for testing trees simulated under various
    \gls{alpha} values. All epidemics had \gls{I} = 1000 infected nodes, on
    \gls{BA} networks of size \gls{N} = 5000 with \gls{m} fixed at 2. Colours
    indicate the number of sampled tips.
  }
  \label{fig:gridkernel}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{gridsearch-alpha-point-estimate.pdf}
  \caption[Marginal estimates of \gls{alpha} obtained with grid search]{
    Marginal estimates of \gls{alpha} obtained with grid search. Training trees
    were simulated on a narrowly spaced grid of \gls{alpha} values, and
    compared to testing trees using the tree kernel. The \gls{alpha} value in
    the grid with the highest median kernel score was taken as the point
    estimate for the testing tree. These point estimates are shown as black
    dots. The dashed line is the identity.
  }
  \label{fig:gridest}
\end{figure}

\subsection{Accuracy of estimates with full ABC}

@@

\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
distribution on the \gls{BA} parameters for one simulation (equivalent plots
for all the simulations can be found in the supplemental materials). \Gls{HPD}
intervals around \gls{alpha} and \gls{I} were narrow relative to the region of
nonzero prior density, whereas the intervals for $m$ and \gls{N} were widely
dispersed. Table~\ref{tab:abchpd} shows point estimates and 95\% \gls{HPD}
intervals averaged over all simulations.

@@

\subsection{Characterization of power-law exponent in Barab\'asi-Albert networks}

@@

>>>>

\subsection{Analysis of \acrlong{BA} model}

<<setup, include=FALSE>>=
    library(netabc)
    library(Hmisc)
@

\subsubsection*{Classifiers for parameters based on tree shape}

@@

    pp <- function (p) {
        if (p < 1e-5) {
            "<10^{-5}"
        } else {
            latexSN(round(p, -floor(log10(p))))
        }
    }

@@

    alpha.sackin.cor <- alpha.stats[,cor.test(alpha, sackin, method="spearman")]
    I.sackin.cor <- I.stats[,cor.test(I, sackin, method="spearman")]
    m.sackin.cor <- m.stats[,cor.test(m, sackin, method="spearman")]
    N.sackin.cor <- N.stats[,cor.test(N, sackin, method="spearman")]
    stopifnot(alpha.sackin.cor$p.value < 1e-5)

    alpha.ratio.cor <- alpha.stats[,cor.test(int.tip.ratio, alpha)]
    I.ratio.cor <- I.stats[,cor.test(int.tip.ratio, I)]
    m.ratio.cor <- m.stats[,cor.test(int.tip.ratio, m)]
    N.ratio.cor <- N.stats[,cor.test(int.tip.ratio, N)]
    stopifnot(alpha.ratio.cor$p.value < 1e-5)
    stopifnot(I.ratio.cor$p.value < 1e-5)
    stopifnot(m.ratio.cor$p.value < 1e-5)
    stopifnot(N.ratio.cor$p.value < 1e-5)
@

Trees simulated under different values of \gls{alpha} were visibly quite
distinct (\cref{fig:alphatrees}). In particular, higher values of \gls{alpha}
produce networks with a small number of highly connected nodes which, once
infected, are likely to transmit to many other nodes. This results in a more
unbalanced, ladder-like structure in the phylogeny, compared to networks with
lower \gls{alpha} values. None of the other three parameters produced trees
which were as easily distinguished from each other
(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}).
Sackin's index, which measures tree imbalance, was significantly correlated with
all four parameters
    (for $\alpha$, $I$, $m$, and $N$ respectively: Spearman's rho =
     \Sexpr{round(alpha.sackin.cor$estimate, 2)},
     \Sexpr{round(I.sackin.cor$estimate, 2)},
     \Sexpr{round(m.sackin.cor$estimate, 2)},
     \Sexpr{round(N.sackin.cor$estimate, 2)};
     $p$-values
     $\Sexpr{pp(alpha.sackin.cor$p.value)}$,
     $\Sexpr{pp(I.sackin.cor$p.value)}$,
     $\Sexpr{pp(m.sackin.cor$p.value)}$,
     $\Sexpr{pp(N.sackin.cor$p.value)}$)
The ratio of internal to terminal branch lengths was negatively correlated with
\gls{alpha} and \gls{I}, and positively corelated with \gls{N} and \gls{m}
  (Spearman's rho
    \Sexpr{round(alpha.ratio.cor$estimate, 2)},
    \Sexpr{round(I.ratio.cor$estimate, 2)},
    \Sexpr{round(m.ratio.cor$estimate, 2)},
    \Sexpr{round(N.ratio.cor$estimate, 2)};
  $p$-values
    $\Sexpr{latexSN(round(alpha.ratio.cor$p.value, -floor(log10(alpha.ratio.cor$p.value))))}$,
    $\Sexpr{latexSN(round(I.ratio.cor$p.value, -floor(log10(I.ratio.cor$p.value))))}$,
    $\Sexpr{latexSN(round(m.ratio.cor$p.value, -floor(log10(m.ratio.cor$p.value))))}$,
    $\Sexpr{latexSN(round(N.ratio.cor$p.value, -floor(log10(N.ratio.cor$p.value))))}$).

@@

\subsubsection*{Marginal parameter estimates with grid search}

<<gridsearch, include=FALSE>>=
    d <- setDT(collect.data("../../simulations/aggregates/gridsearch/*"))
    d[,error := abs(true_value - point.est)]

    # correlation between number of tips and error
    tip.cor <- d[,cor.test(error, tips, method="spearman")[c("p.value", "estimate")], by=parameter]
    setkey(tip.cor, parameter)

    # CI takes up >90% of grid
    d[parameter == "alpha", all(upper - lower >= 1.8)]
    d[parameter == "I", all(upper - lower >= 4050)]
    d[parameter == "N", all(upper - lower >= 12555)]
    d[parameter == "m", all(upper - lower >= 5)]

    # correlation between value and error
    value.cor <- d[,cor.test(error, true_value)[c("p.value", "estimate")], by=parameter]
    setkey(value.cor, parameter)
    stopifnot(value.cor["I", p.value] >= 0.05)
@

The accuracy of grid search estimates largely paralleled that of the \gls{kSVM}
classifiers. \Cref{fig:gridest} shows point estimates and 95\% highest density
intervals for each of the \gls{BA} parameters, for one replicate experiment
with 500-tip trees. For all parameters except $m$, the error of point estimates
was negatively correlated with the number of sampled tips in the tree (for
\gls{alpha}, \gls{I}, and \gls{N} respectively: Spearman's $\rho$ = 
    \Sexpr{tip.cor["alpha", round(estimate, 2)]},
    \Sexpr{tip.cor["I", round(estimate, 2)]},
    \Sexpr{tip.cor["N", round(estimate, 2)]};
$p$-value = 
    $\Sexpr{tip.cor["alpha", latexSN(round(p.value, -floor(log10(p.value))))]}$,
    $\Sexpr{tip.cor["I", latexSN(round(p.value, -floor(log10(p.value))))]}$,
    $\Sexpr{tip.cor["N", latexSN(round(p.value, -floor(log10(p.value))))]})$.
The highest density intervals obtained for all parameters were extremely wide,
occcupying $>$90\% of the grid in all cases.

The \gls{alpha} parameter was the most accurately estimated, with point
estimates having an average deviation of 
    \Sexpr{d[parameter == "alpha", round(mean(error), 2)]}
from the true value, on a grid from 0 to 2. The error was negatively correlated
with the true value of \gls{alpha} 
    (Spearman's $\rho$ = \Sexpr{value.cor["alpha", round(estimate, 2)]},
     $p = \Sexpr{value.cor["alpha", latexSN(round(p.value, -floor(log10(p.value))))]}$),
although the relationship was clearly nonlinear (\cref{fig:gridest,fig:gridalpha}).
The accuracy was highest for the test value \gls{alpha} = 1.25
    (mean error \Sexpr{d[parameter == "alpha" & true_value == 1.25, round(mean(error), 2)]})
which exhibited markedly different behaviour than the other values in terms of
the distribution of kernel scores along the grid (\cref{fig:gridalpha}). In
particular, there was a very pronounced peak in scores around the true value,
in contrast to most other values where the scores were flat around the true
value. The peak was also observed to a lesser degree for \gls{alpha} = 1. The
average absolute error of the point estimates for \gls{I} was 
    \Sexpr{d[parameter == "I", round(mean(error))]},
on a grid of 500 to 5000, and the errors were not significantly correlated with
the true value of \gls{I}. Kernel score distributions for all test values
exhibited the same rounded shape (\cref{fig:gridI}).

For the other two parameters, \gls{m} and \gls{N}, accuracy was much poorer
(\cref{fig:gridest,fig:gridm,fig:gridN}). The average error for \gls{m} was
    \Sexpr{d[parameter == "m", round(mean(error), 2)]},
on a grid from 1 to 6; this error was positively correlated with increasing
\gls{m}
    (Spearman's $\rho$ = \Sexpr{value.cor["m", round(estimate, 2)]},
     $p = \Sexpr{value.cor["m", latexSN(round(p.value, -floor(log10(p.value))))]}$).
This positive correlation was apparently due to the much lower error for $m =
1$ than for the other $m$ values (\cref{fig:gridm})
    (mean errors \Sexpr{d[parameter == "m" & true_value == 1, mean(error)]} for
     $m = 1$ vs. \Sexpr{d[parameter == "m" & true_value > 1, mean(error)]} for
     $m > 1$).
The value $m = 1$ causes the network to take on a distinct shape relative to
higher $m$ values, namely a tree (\textit{i.e.} there are no cycles,
\cref{subsec:treeshape}). The average error for \gls{N} was 
    \Sexpr{d[parameter == "N", round(mean(error))]},
on a grid from 1000 to 15000, and was positively correlated with the true value
of \gls{N}
    (Spearman's $\rho$ = \Sexpr{value.cor["N", round(estimate, 2)]},
     $p = \Sexpr{value.cor["N", latexSN(round(p.value, -floor(log10(p.value))))]}$).

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{gridsearch-example}
  \caption[Grid search estimates of \gls{BA} model parameters]{Point estimates
      and 95\% highest density intervals for each \gls{BA} model parameter,
      obtained using grid search. Networks and transmission trees were
      simulated over a grid of values for each parameter while holding the
      others fixed. For a subset of the grid values ($x$-axis), test networks
      and trees were created and compared to each tree on the grid using the
      tree kernel. The kernel scores along the grid were normalized to resemble
      a probability distribution, from which the mode and highest density
      interval were calculated. Shown values correspond to one replicate
      experiment, with trees of size 500.
  } 
  \label{fig:gridest}
\end{figure}

\subsubsection*{Joint parameter estimates with kernel-ABC}

@@

\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
distribution on the \gls{BA} parameters for one simulation (equivalent plots
for all the simulations can be found in the supplemental materials). \Gls{HPD}
intervals around \gls{alpha} and \gls{I} were narrow relative to the region of
nonzero prior density, whereas the intervals for $m$ and \gls{N} were widely
dispersed. \Cref{tab:abchpd} shows point estimates and 95\% \gls{HPD}
intervals averaged over all simulations.

@@

\subsubsection*{Effect of parameters on power-law exponent}

@@

\subsection{Application to HIV data}

<<realdata, include=FALSE>>=
  d <- fread(Sys.glob("../../simulations/aggregates/hpd/*"))
  map <- dcast.data.table(d, dataset~parameter, value.var="map")
  up <- dcast.data.table(d, dataset~parameter, value.var="hpd.max")
  low <- dcast.data.table(d, dataset~parameter, value.var="hpd.min")
  setkey(map, dataset)
  setkey(up, dataset)
  setkey(low, dataset)
@

We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
and found substantial heterogeneity among the parameter estimates
(\cref{fig:abchpd}). Two of the datasets~\autocite{niculescu2015recent,
wang2015targeting} had estimated $\alpha$ values near unity (MAP estimates
[95\% HPD] 
  \Sexpr{map["niculescu2015", round(alpha, 2)]} 
  [\Sexpr{low["niculescu2015", round(alpha, 2)]} - 
   \Sexpr{up["niculescu2015", round(alpha, 2)]}]
and
  \Sexpr{map["wang2015", round(alpha, 2)]} 
  [\Sexpr{low["wang2015", round(alpha, 2)]} -
   \Sexpr{up["wang2015", round(alpha, 2)]}] respectively). 
Another two datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated
values and wider HPD intervals
  (\Sexpr{map["li2015", round(alpha, 2)]} 
  [\Sexpr{low["li2015", round(alpha, 2)]} - 
  \Sexpr{up["li2015", round(alpha, 2)]}]
and
  \Sexpr{map["cuevas2009", round(alpha, 2)]} 
  [\Sexpr{low["cuevas2009", round(alpha, 2)]} -
   \Sexpr{up["cuevas2009", round(alpha, 2)]}]). 
The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
and a very wide HPD interval
  (\Sexpr{map["novitsky2014", round(alpha, 2)]} 
  [\Sexpr{low["novitsky2014", round(alpha, 2)]} -
   \Sexpr{up["novitsky2014", round(alpha, 2)]}]). 
For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
$I$ were below 2000, with narrow HPD intervals around two of the
datasets
  (\citeauthor{cuevas2009hiv}, \Sexpr{map["cuevas2009", floor(I)]} 
  [\Sexpr{low["cuevas2009", floor(I)]} -
   \Sexpr{up["cuevas2009", floor(I)]}];
   \citeauthor{niculescu2015recent}, \Sexpr{map["niculescu2015", floor(I)]}
  [\Sexpr{low["niculescu2015", floor(I)]} - 
   \Sexpr{up["niculescu2015", floor(I)]}])
and wider intervals around the other two
  (\citeauthor{li2015hiv}, \Sexpr{map["li2015", floor(I)]} 
  [\Sexpr{low["li2015", floor(I)]} -
   \Sexpr{up["li2015", floor(I)]}];
   \citeauthor{wang2015targeting}, \Sexpr{map["wang2015", floor(I)]}
  [\Sexpr{low["wang2015", floor(I)]} - 
   \Sexpr{up["wang2015", floor(I)]}]).
The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
estimated $I$, and HPD interval spanning almost the entire prior region
  (\Sexpr{map["novitsky2014", floor(I)]} 
  [\Sexpr{low["novitsky2014", floor(I)]} -
   \Sexpr{up["novitsky2014", floor(I)]}]).
No information was gleaned about the $m$ parameter, with the HPD interval
occupying the entire prior region for all datasets. The estimates of $N$ were
similarly uninformative, with the exception that the point estimate for the
\citeauthor{wang2015targeting} data was smaller
  (\Sexpr{map["wang2015", floor(N)]})
than the estimates for other datasets
 (average \Sexpr{map[dataset != "wang2015", floor(mean(N))]}).

\begin{figure*}[ht]
  \centering
  \includegraphics{realdata-hpd-bc.pdf}
  \vspace{8pt}
  \caption{
      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
      parameters of the BA network model, fitted to five published HIV datasets
      with kernel-ABC.
  }
  \label{fig:abchpd}
\end{figure*}
